---
title: "Project graphics"
author: "Kaimo Zhang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=4, fig.height=4)
```


```{r}
library(pxR)
library(ggplot2)
library(tidyverse)
ROA16<-as.data.frame(read.px("ROA16.px"))
df1<-subset(ROA16,Statistic=="Killed Casualties (Number)"&Age.Group=="All ages"&Road.User.Type != "All Car users"&Road.User.Type != "All road users")
df2<-subset(ROA16,Statistic=="Injured Cassualties (Number)"&Age.Group=="All ages"&Road.User.Type != "All Car users"&Road.User.Type != "All road users")
df3<-subset(ROA16,Statistic=="Killed Casualties (Number)"&Age.Group!="All ages"&Road.User.Type == "All road users")
df4<-subset(ROA16,Statistic=="Injured Cassualties (Number)"&Age.Group!="All ages"&Road.User.Type == "All road users")
```

Killed Casualties divided by Year & Sex
```{r}
ggplot(data = df1) + geom_col(mapping = aes(x = Year, y=value, fill=Sex))+coord_flip()

```

Killed Casualties divided by Year & Road.User.Type
```{r}
ggplot(data = df1) + geom_col(mapping = aes(x = Year, y=value, fill=Road.User.Type))+coord_flip()
```

Killed Casualties divided by Year & Age.Group
```{r}
ggplot(data = df3) + geom_col(mapping = aes(x = Year, y=value, fill=Age.Group))+coord_flip()
```

Killed Casualties divided by Road.User.Type & Sex
```{r}
ggplot(data = df1) + geom_col(mapping = aes(x = Road.User.Type, y=value, fill=Sex))+coord_flip()
```

Killed Casualties divided by Age.Group & Sex
```{r}
ggplot(data = df3) + geom_col(mapping = aes(x = Age.Group, y=value, fill=Sex))+coord_flip()
```

Injured Cassualties divided by Year & Sex
```{r}
ggplot(data = df2) + geom_col(mapping = aes(x = Year, y=value, fill=Sex))+coord_flip()
```

Injured Cassualties divided by Year & Road.User.Type
```{r}
ggplot(data = df2) + geom_col(mapping = aes(x = Year, y=value, fill=Road.User.Type))+coord_flip()
```

Injured Cassualties divided by Year & Age.Group
```{r}
ggplot(data = df4) + geom_col(mapping = aes(x = Year, y=value, fill=Age.Group))+coord_flip()
```

Injured Cassualties divided by Road.User.Type & Sex
```{r}
ggplot(data = df2) + geom_col(mapping = aes(x = Road.User.Type, y=value, fill=Sex))+coord_flip()
```

Injured Cassualties divided by Age.Group & Sex
```{r}
ggplot(data = df4) + geom_col(mapping = aes(x = Age.Group, y=value, fill=Sex))+coord_flip()
```
