---
title: "R_PROJECT"
author: "AVEEK DAS"
date: "12/6/2020"
output: html_document
---

### Understanding the Road User Type based on Age Group and Gender

```{r}
library(pxR)
library(ggplot2)
library(tidyverse)
df = as.data.frame(read.px("ROA16.px"))
head(df)
```

```{r}

# Filter all the redundant data

df.not.all_killed_injured <- df %>%
  filter(df$Statistic != "All Killed and Injured Casualties (Number)")
  
df.not.all_road_users <- df.not.all_killed_injured %>%  
  filter(df.not.all_killed_injured$Road.User.Type != "All road users")

df.not.all_car_users <- df.not.all_road_users %>%  
  filter(df.not.all_road_users$Road.User.Type != "All Car users")

df.not.all_ages <- df.not.all_car_users %>%  
  filter(df.not.all_car_users$Age.Group != 'All ages')

df <- df.not.all_ages

sum(df$value) #112279
```

```{r}
p<-ggplot(data=df, aes(x=Road.User.Type, y=value)) +
  geom_bar(stat="identity")
p
```


```{r}
ggplot(data=df, aes(x=Road.User.Type, y=value, fill=Statistic), position = "fill") +
  geom_bar(stat="identity")
```

From the above, we can conclude that the Car Drivers having the highest Casualties Count.


#### Driving Casualities by Gender per Year

```{r}
df.female = df[df$Sex == 'Female',]
sum.female <- aggregate(value~Year, df.female, sum)

df.female.car_drivers = df[df$Road.User.Type == "Car drivers" & df$Sex == 'Female',]
sum.female.car_drivers <- aggregate(value~Year, df.female.car_drivers, sum)

prop.female = round(sum.female.car_drivers$value/sum.female$value, 2)
prop.female





df.male = df[df$Sex == 'Male', ]
sum.male <- aggregate(value~Year, df.male, sum)

df.male.car_drivers = df[df$Road.User.Type == "Car drivers" & df$Sex == 'Male', ]
sum.male.car_drivers <- aggregate(value~Year, df.male.car_drivers, sum)

prop.male = round(sum.male.car_drivers$value/sum.male$value, 2)
prop.male

```


### Proportion of Male and Female Car Driving Death Casualties Trend over 14 years.

```{r}
df.killed.female.car_drivers = df[df$Statistic =='Killed Casualties (Number)' & df$Road.User.Type == "Car drivers" & df$Sex == 'Female', ]
sum.killed.female.car_drivers <- aggregate(value~Year, df.killed.female.car_drivers, sum)

df.female.car_drivers = df[df$Road.User.Type == "Car drivers" & df$Sex == 'Female', ]
sum.female.car_drivers <- aggregate(value~Year, df.female.car_drivers, sum)

killed.female.prop = round(sum.killed.female.car_drivers$value/sum.female.car_drivers$value, 2)
killed.female.prop




df.killed.male.car_drivers = df[df$Statistic =='Killed Casualties (Number)' & df$Road.User.Type == "Car drivers" & df$Sex == 'Male', ]
sum.killed.male.car_drivers <- aggregate(value~Year, df.killed.male.car_drivers, sum)

df.male.car_drivers = df[df$Road.User.Type == "Car drivers" & df$Sex == 'Male', ]
sum.male.car_drivers <- aggregate(value~Year, df.male.car_drivers, sum)
killed.male.prop = round(sum.killed.male.car_drivers$value/sum.male.car_drivers$value, 2)
killed.male.prop
```

# Categorical Count of AGe Group WIse Casualties for 14 years.

```{r}
ggplot(data=df, aes(x=Age.Group, y=value, fill=Statistic), position = "fill") +
  geom_bar(stat="identity")
```


# The age group containing 25-34 years has the highest number of casualities.

From the above we can conclude that the Road User Type of Pedal Cyclist has an increasing trend of casualties over the years. However, the proportions of fatalities is minimum.
